---
title: "*Vanessa cardui* DE"
output: html_notebook
---
```{r}
library(topGO)
library(ggplot2)
library(DESeq2)
library(apeglm)
library("clusterProfiler")
library("enrichplot")
library("data.table")
```


# Differntial expression in *Vanessa cardui* adults
Experimental set up was prepared with adult females in two food availability conditions: ad libitum and no resources, to analyze the influence that environmental conditions may have on investment in migration or reproduction in imagines

## Abdomen tissue 
In this section we analyse differential expression in the abdomen tissue


**Preparing data**  

We upload count data from the DESeq2 object crated within the nf-core rnasq pipeline. Pipeline was run for the entire experimental dataset therefore needs to be split into different sets/tissues:

```{r}
load("/Users/dshipilina/GitHub/VanessaExpression/deseq2_qc/deseq2.dds.RData") #loading file
countLR<- dds@assays@data@listData[["counts"]] #reloading count data
colDataLR <- colData(dds) #reloading metadata 

#Renaming columns for clarity
names(colDataLR)[names(colDataLR)=="Group3"]<-"tissue"
names(colDataLR)[names(colDataLR)=="Group1"]<-"condition"


#Checking resulting tables
head(countLR)
head(colDataLR)
```

Pipeline was run for the entire experimental dataset therefore needs to be split into different sets/tissues:
#### Selecting data for abdomen
```{r}
colDataLRab<-colDataLR[colDataLR$tissue == "A", ] #selecting abdomen samples only
#selecting samples with abdomen tissue based on ID 
countLRab<-countLR[,c(1,2,3,4,5,10,11,12,13,14)]
names(colDataLRab)[names(colDataLRab)=="Group3"]<-"tissue"
names(colDataLRab)[names(colDataLRab)=="Group1"]<-"condition"

head(countLRab)
head(colDataLRab)
```

Constructing DESeq2 object, at this point counts are transformed according to the DESeq specific algorithm.
Current hypothsis includes split into treatment groups *only* (~condition), filtering out genes with less then 10 total counts 
```{r}
ddsLRab <- DESeqDataSetFromMatrix(countData=countLRab, colData =colDataLRab, design = ~condition)

#Basic filtering
keep <- rowSums(counts(ddsLRab)) >= 10
ddsLRab <- ddsLRab[keep,]
head(ddsLRab)
```

**DESeq analysis**
Main valuation, calculation of p-values
```{r}
ddsLRab <- DESeq(ddsLRab) #running the main function
resLRab <- results(ddsLRab)
resLRab <- resLRab[order(resLRab$padj),]
head(resLRab)
summary(resLRab)
```

Setting up p-value to 0.05
```{r}
res05 <- results(ddsLRab, alpha=0.05)
summary(res05)
```



Log fold change shrinkage is recommended for visualization and ranking, her we follow main DESEq2 vignette exactly

"Shrinkage of effect size (LFC estimates) is useful for visualization and ranking of genes. To shrink the LFC, we pass the dds object to the function lfcShrink. Below we specify to use the apeglm method for effect size shrinkage (Zhu, Ibrahim, and Love 2018), which improves on the previous estimator.

We provide the dds object and the name or number of the coefficient we want to shrink, where the number refers to the order of the coefficient as it appears in resultsNames(dds)."


```{r}
#head(resultsNames(ddsLRab))
resLFC <- lfcShrink(ddsLRab, coef="condition_LI_vs_AL", type="apeglm")
resLFC
```


Visual exploration of the results

We us first 100 significant genes to build a PCA plot:
```{r}
vsdata <- vst(ddsLRab, blind=FALSE)
plotPCA(vsdata, intgroup=c("Group4"))

head(ddsLRab)
```
#### Including group effect into analysis
Let's define new condition

We create new column with family assignment for head:
"AL_ADF_H_11" "AL_ADF_H_3"  "AL_ADF_H_7"  "AL_ADF_H_9"  "LI_ADF_H_10" "LI_ADF_H_12" "LI_ADF_H_14" "LI_ADF_H_16" "LI_ADF_H_8"
"X18" "X5"  "X4"  "X4"  "Unknown" "Unknown" "Unknown" "Unknown" "Unknown"

For abdomens:
"AL_ADF_A_11" "AL_ADF_A_19" "AL_ADF_A_3"  "AL_ADF_A_7"  "AL_ADF_A_9"  "LI_ADF_A_10" "LI_ADF_A_12" "LI_ADF_A_14" "LI_ADF_A_16" "LI_ADF_A_8"
"X18" "X18" "X5" "X4" "X4" "Unknown" "Unknown" "Unknown" "Unknown" "Unknown" "Unknown"

```{r}
colDataLRab<-colDataLR[colDataLR$tissue == "A", ] #selecting abdomen samples only
#selecting samples with abdomen tissue based on ID 
countLRab<-countLR[,c(1,2,3,4,5,10,11,12,13,14)]
names(colDataLRab)[names(colDataLRab)=="Group3"]<-"tissue"
names(colDataLRab)[names(colDataLRab)=="Group1"]<-"condition"

colDataLRab['family'] <- c("X18","X18","X5","X4","X4","X5","X5","U1","U2","U3")
```


We use function plotMA, that shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points are colored blue if the adjusted p-value is less than 0.1.
```{r}
plotMA(resLRab, ylim=c(-5,5))
```

Checking significant gene candidates:
```{r}
res <- resLRab[order(resLRab$padj),]
head(res, 10)

#Plotting these genes
plotCounts(ddsLRab, gene="Vcard_DToL13592", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL04829", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL13539", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL02600", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL10357", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL13220", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL16018", intgroup="condition")
```

Checking functions of the differentialy expressed genes

```{bash}
#ls
grep "Vcard_DToL13592" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL04829" vcard.func.annot.csv | awk -F "," '{print $1,$14,$15}'
grep "Vcard_DToL13539" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL02600" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL10357" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL13220" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL16018" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
```


```{bash}
grep "Vcard_DToL13592" vcard.func.annot.csv 
grep "Vcard_DToL04829" vcard.func.annot.csv 
grep "Vcard_DToL13539" vcard.func.annot.csv 
grep "Vcard_DToL02600" vcard.func.annot.csv 
grep "Vcard_DToL10357" vcard.func.annot.csv 
grep "Vcard_DToL13220" vcard.func.annot.csv 
grep "Vcard_DToL16018" vcard.func.annot.csv 

```



```{r}
tab <- matrix(c("N/A", "N/A", "Isl-1", "Insulin gene enhancer protein, gene encodes a transcription factor containing two N-terminal LIM domains and one C-terminal homeodomain","A10/OS-D","Insect pheromone-binding family","glipr2",1,1,1,1,1,1,1), ncol=2, byrow=TRUE)
colnames(tab) <- c('GeneName','Function')
rownames(tab) <- c('Vcard_DToL13592',"Vcard_DToL04829","Vcard_DToL13539","Vcard_DToL02600","Vcard_DToL10357","Vcard_DToL13220","Vcard_DToL16018" )
tab <- as.table(tab)

tab

```

 Including group effect
```{r}
ddsLRabf <- DESeqDataSetFromMatrix(countData=countLRab, colData = colDataLRab, design = ~family+condition)
keep <- rowSums(counts(ddsLRabf)) >= 10
ddsLRabf <- ddsLRabf[keep,]
head(ddsLRabf)

ddsLRabf <- DESeq(ddsLRabf)
resLRabf <- results(ddsLRabf)
resLRabf <- resLRabf[order(resLRabf$padj),]
head(resLRabf)
summary(resLRabf)
```
Filtering for lower p-values
```{r}
res05abf <- results(ddsLRabf, alpha=0.05)
summary(res05abf)

#vsdataab <- vst(ddsLRabf, blind=FALSE)
#plotPCA(vsdataab, intgroup=c("condition"))

```

#### Results

Visualize results in 
```{r}
#resLFC <- lfcShrink(ddsLRab, coef="condition_LI_vs_AL", type="ashr")
resLFCa <- lfcShrink(ddsLRab, coef="condition_LI_vs_AL", type="apeglm")
#resLFC

#plotMA(resLFC, ylim=c(-3,3))
plotMA(resLFCa, ylim=c(-4,4))
#plotMA(res05abf)#, ylim=c(-2,2))
```




```{r}
res05abf<-na.omit(res05abf)
res05abftopUp <- res05abf[res05abf$padj <= 0.05 & res05abf$log2FoldChange > 0, ]
res05abftopDown <- res05abf[res05abf$padj <= 0.05 & res05abf$log2FoldChange < 0, ]
res05abftopUp
res05abftopDown

res05abftopUpgenenames <- row.names(res05abftopUp)
res05abftopUpgenenames

res05abftopDowngenenames <- row.names(res05abftopDown)
res05abftopDowngenenames
```


```{r}

#resLFC <- lfcShrink(ddsLRab, coef="condition_LI_vs_AL", type="ashr")
##resLFCa <- lfcShrink(ddsLRabf, coef="condition_LI_vs_AL", type="apeglm")
#resLFC
#reset par
par(mfrow=c(1,1))
# Make a basic volcano plot
with(resLFCa, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-1,1)))

# Add colored points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
with(subset(resLFCa, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(resLFCa, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))


#reset par
par(mfrow=c(1,1))
# Make a basic volcano plot
with(res05abf, plot(log2FoldChange, -log10(pvalue), pch=20, col="grey", main="Abdomen", xlim=c(-6,6), ylim=c(0,20)))

# Add colored points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
#with(subset(res05abf, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(res05abf, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="#FF8C00"))
```

```{r}
summary(resLFCa)
resLFCprep<-na.omit(resLFCa)
resLFCabf <- resLFCprep[resLFCprep$padj <= 0.05 & resLFCprep$log2FoldChange > 2, ]
summary(resLFCabf)
```


#### Candidate genes

```{r}
fun_ann <- read.csv(file ="/Users/dshipilina/GitHub/VanessaExpression/vcard.func.annot.csv",header = FALSE)
names(fun_ann)[names(fun_ann)=="V1"]<-"GeneID"
names(fun_ann)[names(fun_ann)=="V8"]<-"Function"
names(fun_ann)[names(fun_ann)=="V9"]<-"GeneName"
names(fun_ann)[names(fun_ann)=="V21"]<-"Func_Categ"
fun_ann$GeneID<-gsub("-RA","",as.character(fun_ann$GeneID))
functional_annotation<-fun_ann[,c("GeneID","Function","GeneName","Func_Categ")]

functional_annotation

```




```{r}
res05abf<-na.omit(res05abf)
res05abftopUp <- res05abf[res05abf$padj <= 0.05 & res05abf$log2FoldChange > 2, ]
res05abftopDown <- res05abf[res05abf$padj <= 0.05 & res05abf$log2FoldChange < 2, ]
length(row.names(res05abftopUp))
length(row.names(res05abftopDown))


res05abftopUpfun <-functional_annotation[is.element(functional_annotation$GeneID,c(row.names(res05abftopUp))),]
write.csv(res05abftopUpfun,file="cand_genes_adult_ab_topUp.tsv",row.names=T)
res05abftopDownfun <- functional_annotation[is.element(functional_annotation$GeneID,c(row.names(res05abftopDown))),]
write.csv(res05abftopDownfun,file="cand_genes_adult_ab_topDown.tsv",row.names=T)

#res05abftopUpgenenames <- row.names(res05abftopUp)
#res05abftopUpgenenames

#res05abftopDowngenenames <- row.names(res05abftopDown)
#res05abftopDowngenenames

```



```{r}
cand_genes_adult_ab_top <-c(row.names(res05abftopUp),row.names(res05abftopDown))
length(cand_genes_adult_ab_top)
#par(mfrow=c(2,3))
#for (x in cand_genes_adult_ab_top) {
#  plotCounts(ddsLRabf, gene=x, intgroup="condition")
#}


###### PLOTTING CANDIDATE GENES #######
 
par(mfrow=c(1,2))

#plotCounts(ddsLRabf, gene="Vcard_DToL01271", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL05461", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL08950", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL01848", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL05597", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL07986", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL11904", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL12126", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL02554", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL02600", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL13323", intgroup="condition")


#plotCounts(ddsLRabf, gene="Vcard_DToL01271", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL05461", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL08950", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL01848", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL05597", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL07986", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL11904", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL12126", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL02554", intgroup="condition")
#plotCounts(ddsLRabf, gene="Vcard_DToL02600", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL13323", intgroup="condition")

cand_genes_adult_ab_best_list <- c("Vcard_DToL01271", "Vcard_DToL05461","Vcard_DToL08950", "Vcard_DToL01848","Vcard_DToL05597","Vcard_DToL07986", "Vcard_DToL11904", "Vcard_DToL12126","Vcard_DToL02554","Vcard_DToL02600", "Vcard_DToL13323")
cand_genes_adult_ab_best <- functional_annotation[is.element(functional_annotation$GeneID,cand_genes_adult_ab_best_list),]
cand_genes_adult_ab_best
```
```{r}
par(mfrow=c(2,3))

for (x in row.names(res05abf)) {
  plotCounts(ddsLRabf, gene=x, intgroup="condition")
}
```


```{r}
res05abftopDowngenenames
y<-res05abftopDowngenenames
write.csv(y,file="abftopDowngenenames.tsv",row.names=F)
```



```{bash}
tail -n +2 abftopDowngenenames.tsv | sed 's/\"//' | sed 's/\"//' | awk '{print $1"-RA"}' > LRabftopALligenenames.tsv
grep -f LRabftopALligenenames.tsv vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}' | sort


#grep -f LRabftopALligenenames.tsv vcard.func.annot.csv 

#grep "Vcard_DToL01271" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL05461" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL05461" vcard.func.annot.csv

```

```{r}

par(mfrow=c(2,3))
for (x in res05abftopUpgenenames) {
  plotCounts(ddsLRabf, gene=x, intgroup="condition")
}


par(mfrow=c(1,1))

plotCounts(ddsLRabf, gene="Vcard_DToL00065", intgroup="condition")
plotCounts(ddsLRabf, gene="Vcard_DToL11904", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL02971", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL06375", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL09542", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL14898", intgroup="condition")
```


```{r}
res05abftopUpgenenames
y<-res05abftopUpgenenames
write.csv(y,file="abftopUpgenenames.tsv",row.names=F)
```


```{bash}

tail -n +2 abftopUpgenenames.tsv | sed 's/\"//' | sed 's/\"//' | awk '{print $1"-RA"}' > LRabfto.tsv
#grep -f LRabftopalLIgenenames.tsv vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}' | sort

```

```{bash}
#grep 'Vcard_DToL00065-RA' vcard.func.annot.csv
#grep 'Vcard_DToL01847-RA' vcard.func.annot.csv | awk -F "," '{print $1,$15,$16,$17,$28}' | sort
#grep 'Vcard_DToL04177-RA' vcard.func.annot.csv
#Vcard_DToL05345-RA
#grep 'Vcard_DToL05345-RA' vcard.func.annot.csv  
#| awk -F "," '{print $1,$13}' 
#grep 'Vcard_DToL05597-RA' vcard.func.annot.csv | awk -F "," '{print $1,$15,$16,$17,$28,$108}' 
#Vcard_DToL07626-RA
#grep 'Vcard_DToL07626-RA' vcard.func.annot.csv | awk -F "," '{print $1,$15,$16,$17,$28,$108}'
#Vcard_DToL08808-RA
grep 'Vcard_DToL08808-RA' vcard.func.annot.csv | awk -F "," '{print $1,$15,$16,$17,$28,$108}'
```



## Head

### Data set up
```{r}
colDataLRhe<-colDataLR[colDataLR$tissue == "H", ] #selecting abdomen samples only
#selecting samples with abdomen tissue based on ID 
countLRhe<-countLR[,-c(1,2,3,4,5,10,11,12,13,14)]
names(colDataLRhe)[names(colDataLRhe)=="Group3"]<-"tissue"
names(colDataLRhe)[names(colDataLRhe)=="Group1"]<-"condition"

colDataLRhe['family'] <- c("X18","X5","X4","X4","Unknown","Unknown","Unknown","Unknown","Unknown")

head(countLRhe)
head(colDataLRhe)
```

Constructing DESeq2 object, at this point counts are transformed according to the DESeq specific algorithm.
Current hypothsis includes split into treatment groups *only* (~condition), filtering out genes with less then 10 total counts 

```{r}
ddsLRhe <- DESeqDataSetFromMatrix(countData=countLRhe, colData =colDataLRhe, design = ~condition)

#Basic filtering
keep <- rowSums(counts(ddsLRhe)) >= 10
ddsLRab <- ddsLRhe[keep,]
head(ddsLRhe)
```

**DESeq analysis**
Main valuation, calculation of p-values
```{r}
ddsLRhe <- DESeq(ddsLRhe) #running the main function
resLRhe <- results(ddsLRhe)
resLRhe <- resLRhe[order(resLRhe$padj),]
head(resLRhe)
summary(resLRhe)
```

Visual exploration of the results

We us first 100 significant genes to build a PCA plot:
```{r}
vsdata <- vst(ddsLRhe, blind=FALSE)
plotPCA(vsdata, intgroup=c("family"))
```

Redefining the families:
```{r}
colDataLRhe<-colDataLR[colDataLR$tissue == "H", ] #selecting abdomen samples only
#selecting samples with abdomen tissue based on ID 
countLRhe<-countLR[,-c(1,2,3,4,5,10,11,12,13,14)]
names(colDataLRhe)[names(colDataLRhe)=="Group3"]<-"tissue"
names(colDataLRhe)[names(colDataLRhe)=="Group1"]<-"condition"

colDataLRhe['family'] <- c("X18","X5","X4","X4","Unknown","Unknown","Unknown","Unknown","Unknown")
colDataLRhe['family2'] <- c("X18","X5","X4","X4","X5","X5","U1","U2","U3")
head(countLRhe)
head(colDataLRhe)
```

Redesigning the experiment:

```{r}
ddsLRhef2 <- DESeqDataSetFromMatrix(countData=countLRhe, colData = colDataLRhe, design = ~family2+condition)
keep <- rowSums(counts(ddsLRhef2)) >= 10
ddsLRhef2 <- ddsLRhef2[keep,]
head(ddsLRhef2)

#vsdata <- vst(ddsLRhef2, blind=FALSE)
#plotPCA(vsdata, intgroup=c("family2"))
```

**DESeq analysis**
Main valuation, calculation of p-values
```{r}
ddsLRhef2 <- DESeq(ddsLRhef2) #running the main function
resLRhef2 <- results(ddsLRhef2)
resLRhef2 <- resLRhef2[order(resLRhef2$padj),]
head(resLRhef2)
summary(resLRhef2)
```

Visual exploration of the results

We use first 100 significant genes to build a PCA plot:
```{r}
vsdata <- vst(ddsLRhef2, blind=FALSE)
plotPCA(vsdata, intgroup=c("condition"))
```

### Exploring results

```{r}
plotMA(resLRhef2, ylim=c(-10,10))
```

```{r}
#reset par
par(mfrow=c(1,1))
# Make a basic volcano plot
with(resLRhef2, plot(log2FoldChange, -log10(pvalue), pch=20, col="grey", main="Head",  xlim=c(-6,6), ylim=c(0,20)))

# Add colored points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
#with(subset(res05abf, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(resLRhef2, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="#FF8C00"))
```


```{r}
res <- resLRhef2[order(resLRhef2$padj),]
head(res, 10)

res2 <- resLRhef2[order(resLRhef2$log2FoldChange),]
head(res2, 10)
```

```{r}
par(mfrow=c(2,3))

#Plotting these genes
plotCounts(ddsLRab, gene="Vcard_DToL15624", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL15622", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL13698", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL15624", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL15622", intgroup="condition")
plotCounts(ddsLRab, gene="Vcard_DToL13698", intgroup="condition")


```

```{bash}
#ls
grep "Vcard_DToL15624" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL15622" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL15622" vcard.func.annot.csv
grep "Vcard_DToL13698" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
grep "Vcard_DToL03156" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL10357" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL13220" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
#grep "Vcard_DToL16018" vcard.func.annot.csv | awk -F "," '{print $1,$15,$16}'
```



Setting up p-value to 0.05
```{r}
res05hef2 <- results(ddsLRhef2, alpha=0.05)
summary(res05hef2)
```


```{r}
#resLFC <- lfcShrink(ddsLRab, coef="condition_LI_vs_AL", type="ashr")
#resLFCh <- lfcShrink(ddsLRhef2, coef="condition_LI_vs_AL", type="apeglm")
#resLFC

#plotMA(resLFC, ylim=c(-3,3))
plotMA(resLFCh, ylim=c(-4,4))
#plotMA(res05abf)#, ylim=c(-2,2))
```

#### Candidate genes
Selecting upregulated genes of interest based on arbitrary (stringent) criteria
```{r}
#res05hef2[order(res05hef2$padj),]
res05hef2<-na.omit(res05hef2)
res05hef2topUp <- res05hef2[res05hef2$padj <= 0.05 & res05hef2$log2FoldChange > 2, ]
res05hef2topDown <- res05hef2[res05hef2$padj <= 0.05 & res05hef2$log2FoldChange < 2, ]
length(row.names(res05hef2topUp))
length(row.names(res05hef2topDown))

res05hef2topUpgenenames <- row.names(res05hef2topUp)
res05hef2topUpgenenames

res05hef2topDowngenenames <- row.names(res05hef2topDown)
res05hef2topDowngenenames


res05hef2topUpfun <-functional_annotation[is.element(functional_annotation$GeneID,c(row.names(res05hef2topUp))),]
write.csv(res05hef2topUpfun,file="cand_genes_adult_he_topUp.tsv",row.names=T)
res05hef2topDownfun <- functional_annotation[is.element(functional_annotation$GeneID,c(row.names(res05hef2topDown))),]
write.csv(res05hef2topDownfun,file="cand_genes_adult_he_topDown.tsv",row.names=T)

```


```{r}
cand_genes_adult_he_top <-c(row.names(res05hef2topUp),row.names(res05hef2topDown))
length(cand_genes_adult_he_top)
#par(mfrow=c(2,3))
#for (x in cand_genes_adult_he_top) {
#  plotCounts(ddsLRhef2, gene=x, intgroup="condition")
#}


###### PLOTTING CANDIDATE GENES #######
 
par(mfrow=c(1,2))

#plotCounts(ddsLRhef2, gene="Vcard_DToL07064", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL09601", intgroup="condition")
plotCounts(ddsLRhef2, gene="Vcard_DToL01495", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL01126", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL02425", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL02971", intgroup="condition")
plotCounts(ddsLRhef2, gene="Vcard_DToL04426", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL05110", intgroup="condition")
plotCounts(ddsLRhef2, gene="Vcard_DToL06375", intgroup="condition")
plotCounts(ddsLRhef2, gene="Vcard_DToL09542", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL11161", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL13323", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL13698", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL14898", intgroup="condition")
#plotCounts(ddsLRhef2, gene="Vcard_DToL17591", intgroup="condition")



cand_genes_adult_he_best_list <- c("Vcard_DToL04426","Vcard_DToL01495","Vcard_DToL06375","Vcard_DToL09542")
cand_genes_adult_he_best <- functional_annotation[is.element(functional_annotation$GeneID,cand_genes_adult_he_best_list),]
cand_genes_adult_he_best
```




## Testing overlaps between gene sets 

```{r}
length(row.names(res05hef2topDown))
length(row.names(res05hef2topUp))
length(row.names(res05abftopDown))
length(row.names(res05abftopUp))

# "Downregulated in heads and abdomens with limited resource"
intersect(row.names(res05hef2topDown),row.names(res05abftopDown))
# Upregulated in heads and abdomens with limited resource
intersect(row.names(res05abftopUp),row.names(res05hef2topUp))
# Downregulated in heads, upregulated in abdomens with limited resource
intersect(row.names(res05abftopUp),row.names(res05hef2topDown))
#Upregulated in heads, downregulated in abdomens with limited resource
intersect(row.names(res05abftopDown),row.names(res05hef2topUp))
#res05hef2topUp
```


```{bash}

echo "Downregulated in heads and abdomens with limited resource"
comm -12 cand_genes_adult_ab_topDown.tsv cand_genes_adult_he_topDown.tsv

echo "Upregulated in heads and abdomens with limited resource"
comm -12 cand_genes_adult_ab_topUp.tsv cand_genes_adult_he_topUp.tsv

echo "Downregulated in heads, upregulated in abdomens with limited resource"
comm -12 cand_genes_adult_ab_topUp.tsv cand_genes_adult_he_topDown.tsv

echo "Upregulated in heads, downregulated in abdomens with limited resource"
comm -12 cand_genes_adult_ab_topDown.tsv cand_genes_adult_he_topUp.tsv

```

```{bash}
awk '{print $1}' CandidateGenesExpandedOrthogroups.txt > CandidateGenesExpandedOrthogroups.tsv

echo "Downregulated in heads limited resource"
comm -12 LRhetopDowngenenames.tsv CandidateGenesExpandedOrthogroups.tsv

echo "Upregulated in heads limited resource"
comm -12 LRhetopUpgenenames.tsv CandidateGenesExpandedOrthogroups.tsv

echo "Downregulated in abdomens limited resource"
comm -12 LRabtopDowngenenames.tsv CandidateGenesExpandedOrthogroups.tsv

echo "Upregulated in abdomens limited resource"
comm -12 LRabtopUpgenenames.tsv CandidateGenesExpandedOrthogroups.tsv
```



## TopGO analysis

```{r}
library(topGO)
library(ggplot2)
```

Creating topGO object
```{r}
#create an object with all gene identifiers and corresponding GO-terms
#readMappings(file, sep = "\t", IDsep = ",")
geneID2GO <- readMappings(file="Vanessa_TopGo_reference_base_filtered.tsv", sep = "\t", IDsep = ",")
str(head(geneID2GO))

#gene universe, vector of gene names
geneNames <- names(geneID2GO)
head(geneNames)

#read the list of genes of interest res05hef2topDowngenenames
#x<-res05hef2topDowngenenames
#write.csv(x,file="x.tsv",row.names=F)
```

Prepping file with all genes of interest

```{r}
#read the list of genes of interest res05hef2topDowngenenames
y<-res05hef2topDowngenenames
write.csv(y,file="hef2topDowngenenames.tsv",row.names=F)

x<-res05hef2topUpgenenames
write.csv(x,file="hef2topUpgenenames.tsv",row.names=F)
```

```{r}
cand_adult_all <- c(row.names(res05hef2topUp),row.names(res05hef2topDown),row.names(res05abftopDown),row.names(res05abftopUp))
#write.csv(cand_adult_all,file="cand_genes_adult_topGO.tsv",row.names = F)
```


```{bash}
#tail -n +2 hef2topDowngenenames.tsv > hef2topDowngenename.tsv
#tail -n +2 hef2topUpgenenames.tsv > hef2topUpgenename.tsv

#cat hef2topDowngenename.tsv hef2topUpgenename.tsv > LRhefgenename.tsv

tail -n +2 cand_genes_adult_topGO.tsv | sed 's/\"//' | sed 's/\"//' | awk '{print $1"-RA"}' > cand_genes_adult_topGOclean.tsv
```


```{r}
candidate_genes <- read.table("cand_genes_adult_topGOclean.tsv", header = F)
colnames(candidate_genes) <- "geneid"
#cand_adult_all
#head(candidate_genes)

candidate_genes <- as.character(candidate_genes$geneid)
geneList <- factor(as.integer(geneNames %in% candidate_genes))
names(geneList) <- geneNames
head(geneList)

##Biological function
#create the topGOdata object
GO_dataAdult_BP <- new("topGOdata", 
               ontology="BP", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)
```



```{r}
candidate_genes <- read.table("resLTR105filtr.tsv", header = F)
colnames(candidate_genes) <- "geneid"
#head(candidate_genes)

candidate_genes <- as.character(candidate_genes$geneid)
geneList <- factor(as.integer(geneNames %in% candidate_genes))
names(geneList) <- geneNames
#head(geneList)

##Biological function
#create the topGOdata object
GO_dataLTR_BP <- new("topGOdata", 
               ontology="BP", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)
```

```{r}
#read the list of genes of interest 
#"when only a list of interesting genes is provided, 
#the user can use only tests statistics that are based on gene counts, 
#like Fisher’s exact test, Z score and alike."
##Algorithms:
#classic uses all GO terms separately
#elim eliminates genes from parentGOterms if child is more specific (bottom up analysis)
#weight trying to determin if GOterm is better representing the list of interesting genes (is more enriched) than any other node from its neighborhood
#parentChild 

Dasha_topGO <- function(gene_list){

candidate_genes <- read.table(gene_list, header = F)
colnames(candidate_genes) <- "geneid"
#head(candidate_genes)

candidate_genes <- as.character(candidate_genes$geneid)
geneList <- factor(as.integer(geneNames %in% candidate_genes))
names(geneList) <- geneNames
#head(geneList)

##Biological function
#create the topGOdata object
GO_data_BP <- new("topGOdata", 
               ontology="BP", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

BP_resultFisher_weight01 <- runTest(GO_data_BP, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_BP)
allRes_BP <- GenTable(GO_data_BP, 
                   weight01Fisher = BP_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

#multiple test correction, method Benjamini-Hochberg
allRes_BP$p_adj <- p.adjust(allRes_BP$weight01Fisher, method = "fdr")
allRes_BP$GO_class <- "BP"

###Molecular function
#create the topGOdata object
GO_data_MF <- new("topGOdata", 
               ontology="MF", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

MF_resultFisher_weight01 <- runTest(GO_data_MF, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_MF)
allRes_MF <- GenTable(GO_data_MF, 
                   weight01Fisher = MF_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_MF$weight01Fisher <- as.numeric(allRes_MF$weight01Fisher)

#multiple test correction, method Benjamini-Hochberg
allRes_MF$p_adj <- p.adjust(allRes_MF$weight01Fisher, method = "fdr")
allRes_MF$GO_class <- "MF"

###Cellular compartment
#create the topGOdata object
GO_data_CC <- new("topGOdata", 
               ontology="CC", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)

CC_resultFisher_weight01 <- runTest(GO_data_CC, algorithm = "weight01", statistic = "fisher")

allGO = usedGO(object = GO_data_CC)
allRes_CC <- GenTable(GO_data_CC, 
                   weight01Fisher = CC_resultFisher_weight01,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_CC$weight01Fisher <- as.numeric(allRes_CC$weight01Fisher)
#multiple test correction, method Benjamini-Hochberg
allRes_CC$p_adj <- p.adjust(allRes_CC$weight01Fisher, method = "fdr")
allRes_CC$GO_class <- "CC"

#assembling all the data
allRes_all_comb <- rbind(allRes_BP[allRes_BP$weight01Fisher<=0.05, ], allRes_CC[allRes_CC$weight01Fisher<=0.05, ], allRes_MF[allRes_MF$weight01Fisher<=0.05, ])

return(allRes_all_comb)

}
```


```{r}
allRes_all_comb<-Dasha_topGO("LRhef2genenames.tsv")
```



```{r}
adul_resultFisher_weight01 <- runTest(GO_dataAdult_BP, algorithm = "weight01", statistic = "fisher")

#BP_resultFisher_classic <- runTest(GO_dataLTR_BP, algorithm = "classic", statistic = "fisher")
#BP_resultFisher_classic

#BP_resultFisher_parentChild <- runTest(GO_dataLTR_BP, algorithm = "parentChild", statistic = "fisher")
#BP_resultFisher_parentChild

#BP_resultFisher_elim <- runTest(GO_dataLTR_BP, algorithm = "elim", statistic = "fisher")
#BP_resultFisher_elim

allGO = usedGO(object = GO_dataAdult_BP)
allResAdult_BP <- GenTable(GO_dataAdult_BP, 
                   weight01Fisher = BP_resultFisher_weight01, 
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allResAdult_BP$weight01Fisher <- as.numeric(allResAdult_BP$weight01Fisher)
allResAdult_BP[allResAdult_BP$weight01Fisher<=0.05, ]

#allResLTR_BP$parentChFisher <- as.numeric(allResLTR_BP$parentChFisher)
#allResLTR_BP[allResLTR_BP$parentChFisher<=0.05, ]

#allResLTR_BP$elimFisher <- as.numeric(allResLTR_BP$elimFisher)
#allResLTR_BP[allResLTR_BP$elimFisher<=0.05, ]


```


Multiple test correction
```{r}
allResAdult_BP$p_adj_fdr <- p.adjust(allResAdult_BP$weight01Fisher, method = "fdr")
allResAdult_BP$p_adj_holm <- p.adjust(allResAdult_BP$weight01Fisher, method = "holm")
allResAdult_BP$p_adj_bonferroni <- p.adjust(allResAdult_BP$weight01Fisher, method = "bonferroni")
allResAdult_BP$p_adj_BH <- p.adjust(allResAdult_BP$weight01Fisher, method = "BH")

allResAdult_BP
```

```{r}
topGOAdult_BP <- allResAdult_BP[allResAdult_BP$weight01Fisher<=0.01, ]
nrow(allResAdult_BP[allResAdult_BP$weight01Fisher<=0.01, ])


topGOAdult_BP$GO_descr <- paste(topGOAdult_BP$Term," (",topGOAdult_BP$GO.ID,")", sep = "")

plot_topGoAdult=ggplot(topGOAdult_BP, aes(x=reorder(GO_descr, -Significant), y=Significant)) + geom_bar(stat="identity", fill="#512887") + 
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "black"),
        axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 8, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 8, angle = 90, vjust = 0.5),
        legend.position = "top",
        axis.ticks = element_line(colour = "black", size = 0.1))

ggsave(plot = plot_topGoAdult, "topGOAdult_BP.pdf")
plot_topGoAdult
```


## KEGG enrichment test

```{r}

### KEGG data prep

# get the necessary information from the eggNOG output file
fun_ann2 <- read.csv(file ="/Users/dshipilina/GitHub/VanessaExpression/vcard.func.annot.csv",header = FALSE)

#get columns 1 (query) and 9 (KO terms)
kegg_data2 <- fun_ann2[c(1,13)]
kegg_data2

# clean up by removing the "ko:" in front of every KO term
kegg_data2$KEGG.KO <- kegg_data2$V13
kegg_data2$Query <- gsub( "-RA", "", as.character(kegg_data2$V1))
kegg_data2

# expand, since some genes/proteins will have multiple assigned KO terms
kegg2 <- data.table(kegg_data2)
kegg2 <- kegg2[, list(KEGG.KO = unlist(strsplit(KEGG.KO, ","))), by = Query]
kegg2_final <- kegg2[,c(2,1)]
kegg2_final

# select the needed columns
#kegg_final <- kegg[,c(2,1)]

# finally you need a list of gene/protein names of interest (only the     # identifiers). Here, this is called protein_ids. It can be a vector or a # column in a data.frame. This can be a list of differentially expressed   # genes or the genes present in a cluster of module that is experimentall # interesting to you.
```


```{r}

#special_genesLRTHDLD <- c(special_genesLTR105,special_genesLTR305)
#genesLRTHDLD <- sort(special_genesLRTHDLD, decreasing = TRUE)
#genesLRTHDLD <- genesLRTHDLD[!duplicated(genesLRTHDLD)]

cand_adult_all_load <- read.csv(file ="/Users/dshipilina/GitHub/VanessaExpression/cand_genes_adult_topGO.tsv",head=FALSE, sep=";")
cand_adult_all_load[-1,]

nrow(kegg2_final[- grep("map", kegg2_final$KEGG.KO),])
kegg_KOterms <- kegg2_final[- grep("map", kegg2_final$KEGG.KO),]

#,header = FALSE)
#write.csv(cand_adult_all,file="cand_genes_adult_topGO.tsv",row.names = F)

enr_resAdult <- enricher(cand_adult_all_load[-1,], TERM2GENE=kegg_KOterms, pvalueCutoff = 0.05, pAdjustMethod = "fdr", qvalueCutoff = 0.2, minGSSize = 1)
enr_resAdult$ID
#enr_res2$ID

# write the results of the analysis
write.table(enr_resAdult, file = "enr_resAdult.csv") # write table

# make a simple dot plot of the results and save it as a .tiff file
tiff(file="KO_dotplotAdult.tiff",width=6, height=4, units="in", res=1200)
dotplot(enr_resAdult, showCategory=30)
dev.off()
```
