---
title: "topGO_Vanessa_cardui_gains"
author: "KN"
date: "14/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene ontology enrichment test for genes gained in Vanessa cardui
Input
1. gene-universe: all genes annotated in Vanessa cardui
2. genes of interest: the output of BadiRate, genes belonging to gene families with different growth rate in Vanessa cardui compared the others (branch specific model higher fit than global model), specifically gene families with gene gains >0 in V. cardui. The gene gains have been parsed with separate scripts get_best_model_gains.sh and the the gene names in the orthogroups are retrieved with gene_selector_mod.sh


```{r setup2, include=FALSE}
library(topGO)
library(ggplot2)
```

```{r data}

#create an object with all gene identifiers and corresponding GO-terms
#readMappings(file, sep = "\t", IDsep = ",")
geneID2GO <- readMappings(file="../TopGO_Data_Prep/Vanessa_TopGo_reference_base_filtered.tsv", sep = "\t", IDsep = ",")
str(head(geneID2GO))

#gene universe, vector of gene names
geneNames <- names(geneID2GO)
head(geneNames)

#read the list of genes of interest
candidate_genes <- read.table("../../results/sp_gains_bestmodel.txt_genenames.txt", header = F)
colnames(candidate_genes) <- "geneid"
head(candidate_genes)

candidate_genes <- as.character(candidate_genes$geneid)
geneList <- factor(as.integer(geneNames %in% candidate_genes))
names(geneList) <- geneNames
head(geneList)
str(geneList)

#create the topGOdata object for each ontology
GO_data_BP <- new("topGOdata", 
               ontology="BP", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)
GO_data_BP

GO_data_CC <- new("topGOdata", 
               ontology="CC", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)
GO_data_CC

GO_data_MF <- new("topGOdata", 
               ontology="MF", 
               allGenes=geneList, 
               annot=annFUN.gene2GO, 
               gene2GO=geneID2GO, 
               nodeSize=1)
GO_data_MF



```


```{r test}
#"when only a list of interesting genes is provided, 
#the user can use only tests statistics that are based on gene counts, 
#like Fisher’s exact test, Z score and alike."
##Algorithms:
#classic uses all GO terms separately
#elim eliminates genes from parentGOterms if child is more specific (bottom up analysis)
#weight trying to determin if GOterm is better representing the list of interesting genes (is more enriched) than any other node from its neighborhood
#parentChild 

#biological process
BP_resultFisher_weight01 <- runTest(GO_data_BP, algorithm = "weight01", statistic = "fisher")
BP_resultFisher_weight01

BP_resultFisher_classic <- runTest(GO_data_BP, algorithm = "classic", statistic = "fisher")
BP_resultFisher_classic

BP_resultFisher_parentChild <- runTest(GO_data_BP, algorithm = "parentChild", statistic = "fisher")
BP_resultFisher_parentChild

BP_resultFisher_elim <- runTest(GO_data_BP, algorithm = "elim", statistic = "fisher")
BP_resultFisher_elim



#allGO = usedGO(object = GOdata)
#topNodes = length(allGO) in GenTable() to get a table with all GO:s to do fdr
allGO = usedGO(object = GO_data_BP)
allRes_BP <- GenTable(GO_data_BP, 
                   weight01Fisher = BP_resultFisher_weight01, 
                   classicFisher = BP_resultFisher_classic,
                   parentChFisher=BP_resultFisher_parentChild,
                   elimFisher=BP_resultFisher_elim,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_BP$weight01Fisher <- as.numeric(allRes_BP$weight01Fisher)
allRes_BP[allRes_BP$weight01Fisher<=0.05, ]

allRes_BP$parentChFisher <- as.numeric(allRes_BP$parentChFisher)
allRes_BP[allRes_BP$parentChFisher<=0.05, ]

allRes_BP$elimFisher <- as.numeric(allRes_BP$elimFisher)
allRes_BP[allRes_BP$elimFisher<=0.05, ]



#multiple test correction, method Benjamini-Hochberg
allRes_BP$p_adj <- p.adjust(allRes_BP$weight01Fisher, method = "fdr")
allRes_BP$p_adj_pc <- p.adjust(allRes_BP$parentChFisher, method = "fdr")
allRes_BP$p_adj_elim <- p.adjust(allRes_BP$elimFisher, method = "fdr")

#adj p-value above 0.05
allRes_BP[allRes_BP$p_adj<=0.05, ]
allRes_BP[allRes_BP$p_adj_pc<=0.05, ]
allRes_BP[allRes_BP$p_adj_elim<=0.05, ]

allRes_BP$GO_class <- "BP"



#cellular compartment
CC_resultFisher_weight01 <- runTest(GO_data_CC, algorithm = "weight01", statistic = "fisher")
CC_resultFisher_weight01

CC_resultFisher_classic <- runTest(GO_data_CC, algorithm = "classic", statistic = "fisher")
CC_resultFisher_classic

CC_resultFisher_parentChild <- runTest(GO_data_CC, algorithm = "parentChild", statistic = "fisher")
CC_resultFisher_parentChild

CC_resultFisher_elim <- runTest(GO_data_CC, algorithm = "elim", statistic = "fisher")
CC_resultFisher_elim



#allGO = usedGO(object = GOdata)
#topNodes = length(allGO) in GenTable() to get a table with all GO:s to do fdr
allGO = usedGO(object = GO_data_CC)
allRes_CC <- GenTable(GO_data_CC, 
                   weight01Fisher = CC_resultFisher_weight01, 
                   classicFisher = CC_resultFisher_classic,
                   parentChFisher=CC_resultFisher_parentChild,
                   elimFisher=CC_resultFisher_elim,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_CC$weight01Fisher <- as.numeric(allRes_CC$weight01Fisher)
allRes_CC[allRes_CC$weight01Fisher<=0.05, ]

allRes_CC$parentChFisher <- as.numeric(allRes_CC$parentChFisher)
allRes_CC[allRes_CC$parentChFisher<=0.05, ]

allRes_CC$elimFisher <- as.numeric(allRes_CC$elimFisher)
allRes_CC[allRes_CC$elimFisher<=0.05, ]



#multiple test correction, method Benjamini-Hochberg
allRes_CC$p_adj <- p.adjust(allRes_CC$weight01Fisher, method = "fdr")
allRes_CC$p_adj_pc <- p.adjust(allRes_CC$parentChFisher, method = "fdr")
allRes_CC$p_adj_elim <- p.adjust(allRes_CC$elimFisher, method = "fdr")

#adj p-value above 0.05
allRes_CC[allRes_CC$p_adj<=0.05, ]
allRes_CC[allRes_CC$p_adj_pc<=0.05, ]
allRes_CC[allRes_CC$p_adj_elim<=0.05, ]

allRes_CC$GO_class <- "CC"

#molecular function
MF_resultFisher_weight01 <- runTest(GO_data_MF, algorithm = "weight01", statistic = "fisher")
MF_resultFisher_weight01

MF_resultFisher_classic <- runTest(GO_data_MF, algorithm = "classic", statistic = "fisher")
MF_resultFisher_classic

MF_resultFisher_parentChild <- runTest(GO_data_MF, algorithm = "parentChild", statistic = "fisher")
MF_resultFisher_parentChild

MF_resultFisher_elim <- runTest(GO_data_MF, algorithm = "elim", statistic = "fisher")
MF_resultFisher_elim



#allGO = usedGO(object = GOdata)
#topNodes = length(allGO) in GenTable() to get a table with all GO:s to do fdr
allGO = usedGO(object = GO_data_MF)
allRes_MF <- GenTable(GO_data_MF, 
                   weight01Fisher = MF_resultFisher_weight01, 
                   classicFisher = MF_resultFisher_classic,
                   parentChFisher=MF_resultFisher_parentChild,
                   elimFisher=MF_resultFisher_elim,
                   orderBy = "weight01Fisher", 
                   ranksOf = "weight01Fisher", 
                   topNodes = length(allGO),
                   numChar=1000)

allRes_MF$weight01Fisher <- as.numeric(allRes_MF$weight01Fisher)
allRes_MF[allRes_MF$weight01Fisher<=0.05, ]

allRes_MF$parentChFisher <- as.numeric(allRes_MF$parentChFisher)
allRes_MF[allRes_MF$parentChFisher<=0.05, ]

allRes_MF$elimFisher <- as.numeric(allRes_MF$elimFisher)
allRes_MF[allRes_MF$elimFisher<=0.05, ]



#multiple test correction, method Benjamini-Hochberg
allRes_MF$p_adj <- p.adjust(allRes_MF$weight01Fisher, method = "fdr")
allRes_MF$p_adj_pc <- p.adjust(allRes_MF$parentChFisher, method = "fdr")
allRes_MF$p_adj_elim <- p.adjust(allRes_MF$elimFisher, method = "fdr")

#adj p-value above 0.05
allRes_MF[allRes_MF$p_adj<=0.05, ]
allRes_MF[allRes_MF$p_adj_pc<=0.05, ]
allRes_MF[allRes_MF$p_adj_elim<=0.05, ]

allRes_MF$GO_class <- "MF"

#add gene names per GO (candidate genes)
allRes_BP$genes <- sapply(allRes_BP$GO.ID,
                          function(x)
                            {
                            genes_sel <- genesInTerm(GO_data_BP, x)
                            genes_sel[[1]][genes_sel[[1]] %in% candidate_genes]
                            })
allRes_BP$genes <- vapply(allRes_BP$genes, paste, collapse =",", character(1L))

allRes_CC$genes <- sapply(allRes_CC$GO.ID,
                          function(x)
                            {
                            genes_sel <- genesInTerm(GO_data_CC, x)
                            genes_sel[[1]][genes_sel[[1]] %in% candidate_genes]
                            })

allRes_CC$genes <- vapply(allRes_CC$genes, paste, collapse =",", character(1L))

allRes_MF$genes <- sapply(allRes_MF$GO.ID,
                          function(x)
                            {
                            genes_sel <- genesInTerm(GO_data_MF, x)
                            genes_sel[[1]][genes_sel[[1]] %in% candidate_genes]
                            })

allRes_MF$genes <- vapply(allRes_MF$genes, paste, collapse =",", character(1L))



```

```{r write results, echo=F, eval=F}
write.table(allRes_BP[1:50,], file = "../result_topGO/V_cardui_res_topGO_VC_BP_50.txt")
write.table(allRes_CC[1:50,], file = "../result_topGO/V_cardui_res_topGO_VC_CC_50.txt")
write.table(allRes_MF[1:50,], file = "../result_topGO/V_cardui_res_topGO_VC_MF_50.txt")

write.table(rbind(allRes_BP[allRes_BP$p_adj<=0.05, ], allRes_CC[allRes_CC$p_adj<=0.05, ], allRes_MF[allRes_MF$p_adj<=0.05, ]), file = "../result_topGO/V_cardui_res_topGO_VC_sign_new.txt", row.names = FALSE)

#semicolon sep, without quotes
write.table(rbind(allRes_BP[allRes_BP$p_adj<=0.05, ], allRes_CC[allRes_CC$p_adj<=0.05, ], allRes_MF[allRes_MF$p_adj<=0.05, ]), file = "../result_topGO/V_cardui_res_topGO_VC_sign_new.csv", row.names = FALSE, sep = ";", quote = FALSE)


allRes_BP[allRes_BP$p_adj<=0.05, ]$genes

```


```{r plots_no_fdr}

allRes_all_comb <- rbind(allRes_BP[allRes_BP$weight01Fisher<=0.05, ], allRes_CC[allRes_CC$weight01Fisher<=0.05, ], allRes_MF[allRes_MF$weight01Fisher<=0.05, ])

allRes_all_comb$GO_descr <- paste(allRes_all_comb$Term," (",allRes_all_comb$GO.ID,")", sep = "")

#plot go-terma and number of significant genes
ggplot(allRes_all_comb, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class)) +
  facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  ylab("Number of genes (p-value < 0.01)") +
  scale_fill_discrete(guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 12, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 12, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 12, angle = 90, vjust = 0.5),
        legend.position = "top"
        )




```

```{r plots_fdr}
#(Not needed, can expand the names it the code earlier)use table with adjusted names (manually checked names in https://www.ebi.ac.uk/QuickGO/search/ to get full names of terms

#if not in env
allRes_all_fdr <- read.csv(file = "../result_topGO/V_cardui_res_topGO_VC_sign.csv",header=T, sep = ";")

#if in env
allRes_all_fdr <- rbind(allRes_BP[allRes_BP$p_adj<=0.05, ], allRes_CC[allRes_CC$p_adj<=0.05, ], allRes_MF[allRes_MF$p_adj<=0.05, ])

allRes_all_fdr$GO_descr <- paste(allRes_all_fdr$Term," (",allRes_all_fdr$GO.ID,")", sep = "")


#plot go-terms and number of significant genes

plot_topGo <- ggplot(allRes_all_fdr, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class)) +
  facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  ylab("Number of genes") +
  scale_fill_discrete(type=c("#d19200","#4d3500", "#916400"),guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.ticks = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 10, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 10, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 10, angle = 90, vjust = 0.5),
        legend.position = c(0.5,0.8),
        legend.direction = "horizontal"
        )


ggsave(plot = plot_topGo, "../result_topGO/V_cardui_barplot_brown.pdf")

#ggsave(plot = plot_topGo, "../result_topGO/V_cardui_barplot_brown.png")

###parentChild
allRes_all_fdr_pc <- rbind(allRes_BP[allRes_BP$p_adj_pc<=0.05, ], allRes_CC[allRes_CC$p_adj_pc<=0.05, ], allRes_MF[allRes_MF$p_adj_pc<=0.05, ])

allRes_all_fdr_pc$GO_descr <- paste(allRes_all_fdr_pc$Term," (",allRes_all_fdr_pc$GO.ID,")", sep = "")


#plot go-terms and number of significant genes


plot_topGo <- ggplot(allRes_all_fdr_pc, aes(reorder(GO_descr, -Significant), Significant)) +
    geom_bar(stat = "identity", aes(fill=GO_class)) +
  facet_grid(~GO_class, scales = "free_x", space = "free_x") +
  ylab("Number of genes (parentChild)") +
  scale_fill_discrete(type=c("#d19200","#4d3500", "#916400"),guide = guide_legend(label.position = "top")) +
  theme(panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),
        axis.line = element_line(size = 0.2, colour = "grey"),
        axis.ticks = element_line(size = 0.2, colour = "grey"),
        axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(size = 10, angle = 90, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 10, angle = 90),
        legend.title = element_blank(),
        legend.text = element_text(size = 10, angle = 90, vjust = 0.5),
        legend.position = c(0.5,0.8),
        legend.direction = "horizontal"
        )


ggsave(plot = plot_topGo, "../result_topGO/V_cardui_barplot_pc.pdf")


```